Missing the Point\(\colon\) Non-Convergence in Multiple Imputation Procedures

### 2020-03-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Included plots with new results in manuscript while working in wrong branch. Going to fix it now.
Use rticles package instead of resolving issues in csl myself.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  Royal, times, sageapa]{sagej}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Missing the Point: Detecting Non-Convergence in Multiple Imputation Procedures},
  pdfauthor={Hanne I. Oberman},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

\title{Missing the Point: Detecting Non-Convergence in Multiple Imputation
Procedures}
\author{Hanne I. Oberman}
\date{16-3-2020}

\begin{document}
\maketitle

@@

At some point, any scientist conducting statistical analyses will run
into a missing data problem (Allison 2001). Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results (Van Buuren 2018). A popular
answer to the ubiquitous problem of missing information is to use the
framework of multiple imputation (MI), proposed by Rubin (1987). For the
theoretical foundation of MI, see Rubin (1987). For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~Van Buuren (2018). With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model}. The imputation model can be a
joint model for all variables, or several univariate conditional
distributions (fully conditional specification). FCS is more flexible
and ``avoids the problem of specifying an appropriate joint imputation
distribution and replaces this by the selection of appropriate
univariate conditional distributions''. With FCS, an iterative
algorithmic procedure is used to \ldots{} multiple imputation by chained
equations \ldots{} ``The values for imputing the missing data are drawn
from the posterior predictive distribution of the missing data
conditional on the observed data.'' There is no scientific consensus on
the convergence properties of FCS algorithms (Takahashi 2017). Some
default techniques in \texttt{mice} might not yield converged states at
all (Murray 2018). Therefore, algorithmic convergence should be
monitored carefully.

Convergence cannot be determined for sure, because MICE is an MCMC
method, and convergence of MCMC is not from scalar to a point, but from
one distribution to another. Generated values will vary even after
convergence. Therefore we can only diagnose non-convergence, not
convergence (Hoff 2009). ``a weak diagnostics is better than no
diagnostic at all'' (Cowles and Carlin 1996).

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye (Van Buuren
2018, \(\S\) 6.5.2). Therefore diagnostic evaluation of convergence
would be preferred. However, not known whether MCMC diagnostics are
appropriate for MICE \ldots{} The application of convergence diagnostics
to MI methods has not been systematically studied (Van Buuren 2018). How
can non-convergence be diagnosed? How many iterations are
sufficient/required? What are the effects of non-convergence on
estimates, predictions and inference? What is an appropriate
non-convergence diagnostic? Are the default number of iterations
sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 1;
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

For reasons of brevity, we only focus on the MI algorithm implemented in
the world-leading MI software: the \texttt{R} (R Core Team 2019) package
\texttt{mice} (Van Buuren and Groothuis-Oudshoorn 2011). The convergence
properties of this MI algorithm are investigated through model-based
simulation. The results of this simulation study are guidelines for
assessing convergence of MI algorithms, which will aid applied
researchers in drawing valid inference from incomplete datasets.

@@

The stationarity component of convergence may be evaluated with
autocorrelation (\(AC\); Schafer 1997; Gelman et al. 2013), numeric
standard error (or `MC error'; Geweke 1992), and Raftery and Lewis's
(1991) procedure to determine the effect of trending within chains.

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) (a.k.a. `Gelman-Rubin statistic'; Gelman and
Rubin 1992). With an adapted version of \(\widehat{R}\), proposed by
Vehtari et al. (2019), we might also evaluate the stationarity component
of convergence. This would make \(\widehat{R}\) a general convergence
diagnostic. The application of \(\widehat{R}\) to assess stationarity
has not been thoroughly investigated. Therefore, this study employs both
\(\widehat{R}\) and autocorrelation to investigate convergence, as
recommended by (Cowles and Carlin 1996, 898).

Note that all of these methods evaluate the convergence of univariate
scalar summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). Van Buuren (2018) proposed
to implement multivariable evaluation through eigenvalue decomposition
(MacKay and Mac Kay 2003). This method is outside of the scope of the
current study (\textbf{not anymore!}).

@@

To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
5). Let \(M\) be the total number of chains, \(T\) the number of
iterations per chain, and \(\theta\) the scalar summary of interest
(e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

We can interpret \(\widehat{R}\) as potential scale reduction factor
since it indicates by how much the variance of \(\theta\) could be
shrunken down if an infinite number of iterations per chain would be run
(Gelman and Rubin 1992). This interpretation assumes that chains are
`over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
Over-dispersion implies that the initial values of the chains are `far
away' from the target distribution and each other. When all chains
sample independent of their initial values, the mixing component of
convergence is satisfied, and \(\widehat{R}\)-values will be close to
one. High \(\widehat{R}\)-values thus indicate non-convergence. The
conventionally acceptable threshold for convergence was
\(\widehat{R} < 1.2\) (Gelman and Rubin 1992). More recently, Vehtari et
al. (2019) proposed a more stringent threshold of
\(\widehat{R} < 1.01\).

@@

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain (Lynch 2007, 147). In this study we only consider \(AC\) at lag 1,
i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration
of the same chain.

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods (i.e., average bias, average confidence interval
width, and empirical coverage rate across simulations; Van Buuren 2018,
\(\S\) 2.5.2). That is, there is no baseline measure available to
evaluate performance against.The current practice of visually inspecting
imputation chains for signs of non-mixing or non-stationarity can only
diagnose severely pathological cases of non-convergence. We could also
look at distributional characteristics, and plausibility of imputed
values, see Vink (n.d.). For now, this is outside of the scope of this
study.

Based on an empirical finding (Lacerda, Ardington, and Leibbrandt 2007),
we hypothesize that \(\widehat{R}\) will over-estimate non-convergence
of MI algorithms. The threshold of \(\widehat{R} < 1.01\) will then be
too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed (Brooks and Gelman 1998, 437). In \texttt{mice}, initial
values are chosen randomly from the observed data. Therefore, we cannot
be certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\).
No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic.

@@

The convergence of \texttt{mice} is investigated through model-based
simulation in \texttt{R} (version 3.6.3; R Core Team 2019). The
simulation set-up is summarized in the pseudo-code below. The complete
\texttt{R} script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# pseudo-code of simulation }
\NormalTok{simulate data }
\ControlFlowTok{for}\NormalTok{ (number of simulation runs from }\DecValTok{1}\NormalTok{ to }\DecValTok{1000}\NormalTok{)}
  \ControlFlowTok{for}\NormalTok{ (number of iterations from }\DecValTok{1}\NormalTok{ to }\DecValTok{100}\NormalTok{)}
\NormalTok{    create missingness}
\NormalTok{    impute the missingness}
\NormalTok{    compute convergence diagnostics}
\NormalTok{    perform analysis}
\NormalTok{    pool results}
\NormalTok{    compute simulation diagnostics}
\NormalTok{aggregate convergence and simulation diagnostics}
\end{Highlighting}
\end{Shaded}

@@

\(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
recommendations. \(AC\) is computed with function \texttt{stats::acf()}.

To estimate the quantity of scientific interest, \(Q\), we perform
multiple linear regression on each completed dataset with the function
\texttt{stats::lm()}. We obtain an estimated regression coefficient per
imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
the function \texttt{mice::pool()} to get variance estimates according
to Rubin's (1987) rules, and subsequently implement finite population
pooling conform Vink and Buuren (2014).

@@

Feedback:

\begin{itemize}
\tightlist
\item
  Add more info about figure legends and axes.
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# plot results}
\KeywordTok{load}\NormalTok{(}\StringTok{"../Results/results.Rdata"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% (Neyman 1934), namely 95\%. Similar to
the CIWs, there is some trending across iterations (i.e., the Loess line
is never flat).

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-alli02}{}%
Allison, Paul D. 2001. \emph{Missing Data}. Sage publications.

\leavevmode\hypertarget{ref-broo98}{}%
Brooks, Stephen P., and Andrew Gelman. 1998. ``General Methods for
Monitoring Convergence of Iterative Simulations.'' \emph{Journal of
Computational and Graphical Statistics} 7 (4): 434--55.
\url{https://doi.org/10.1080/10618600.1998.10474787}.

\leavevmode\hypertarget{ref-cowl96}{}%
Cowles, Mary Kathryn, and Bradley P Carlin. 1996. ``Markov Chain Monte
Carlo Convergence Diagnostics: A Comparative Review.'' \emph{Journal of
the American Statistical Association} 91 (434): 883--904.

\leavevmode\hypertarget{ref-gelm13}{}%
Gelman, Andrew, John B. Carlin, Hal S. Stern, David B. Dunson, Aki
Vehtari, and Donald B. Rubin. 2013. \emph{Bayesian Data Analysis}.
Philadelphia, PA, United States: CRC Press LLC.

\leavevmode\hypertarget{ref-gelm92}{}%
Gelman, Andrew, and Donald B. Rubin. 1992. ``Inference from Iterative
Simulation Using Multiple Sequences.'' \emph{Statistical Science} 7 (4):
457--72. \url{https://doi.org/10.1214/ss/1177011136}.

\leavevmode\hypertarget{ref-gewe92}{}%
Geweke, John. 1992. ``Evaluating the Accuracy of Sampling-Based
Approaches to the Calculations of Posterior Moments.'' \emph{Bayesian
Statistics} 4: 641--49.

\leavevmode\hypertarget{ref-hoff09}{}%
Hoff, Peter D. 2009. \emph{A First Course in Bayesian Statistical
Methods}. Springer Texts in Statistics. New York, NY: Springer New York.
\url{https://doi.org/10.1007/978-0-387-92407-6}.

\leavevmode\hypertarget{ref-lace07}{}%
Lacerda, Miguel, Cally Ardington, and Murray Leibbrandt. 2007.
``Sequential Regression Multiple Imputation for Incomplete Multivariate
Data Using Markov Chain Monte Carlo.'' University of Cape Town, South
Africa.

\leavevmode\hypertarget{ref-lync07}{}%
Lynch, Scott M. 2007. \emph{Introduction to Applied Bayesian Statistics
and Estimation for Social Scientists}. Springer Science \& Business
Media.

\leavevmode\hypertarget{ref-mack03}{}%
MacKay, David JC, and David JC Mac Kay. 2003. \emph{Information Theory,
Inference and Learning Algorithms}. Cambridge university press.

\leavevmode\hypertarget{ref-murr18}{}%
Murray, Jared S. 2018. ``Multiple Imputation: A Review of Practical and
Theoretical Findings.'' \emph{Statistical Science} 33 (2): 142--59.
\url{https://doi.org/10.1214/18-STS644}.

\leavevmode\hypertarget{ref-neym34}{}%
Neyman, Jerzy. 1934. ``On the Two Different Aspects of the
Representative Method: The Method of Stratified Sampling and the Method
of Purposive Selection.'' \emph{Journal of the Royal Statistical
Society} 97 (4): 558--625. \url{https://doi.org/10.2307/2342192}.

\leavevmode\hypertarget{ref-raft91}{}%
Raftery, Adrian E, and Steven Lewis. 1991. ``How Many Iterations in the
Gibbs Sampler?'' Washington University Seatle, Department of Statistics,
United States.

\leavevmode\hypertarget{ref-R}{}%
R Core Team. 2019. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria. \url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-rubin87}{}%
Rubin, Donald B. 1987. \emph{Multiple Imputation for Nonresponse in
Surveys}. Wiley Series in Probability and Mathematical Statistics
Applied Probability and Statistics. New York, NY: Wiley.

\leavevmode\hypertarget{ref-scha97}{}%
Schafer, Joseph L. 1997. \emph{Analysis of Incomplete Multivariate
Data}. Chapman; Hall/CRC.

\leavevmode\hypertarget{ref-taka17}{}%
Takahashi, Masayoshi. 2017. ``Statistical Inference in Missing Data by
MCMC and Non-MCMC Multiple Imputation Algorithms: Assessing the Effects
of Between-Imputation Iterations.'' \emph{Data Science Journal} 16
(July): 37. \url{https://doi.org/10.5334/dsj-2017-037}.

\leavevmode\hypertarget{ref-buur18}{}%
Van Buuren, Stef. 2018. \emph{Flexible Imputation of Missing Data}.
Chapman; Hall/CRC.

\leavevmode\hypertarget{ref-mice}{}%
Van Buuren, Stef, and Karin Groothuis-Oudshoorn. 2011. ``Mice:
Multivariate Imputation by Chained Equations in R.'' \emph{Journal of
Statistical Software} 45 (1): 1--67.
\url{https://doi.org/10.18637/jss.v045.i03}.

\leavevmode\hypertarget{ref-veht19}{}%
Vehtari, Aki, Andrew Gelman, Daniel Simpson, Bob Carpenter, and
Paul-Christian Bürkner. 2019. ``Rank-Normalization, Folding, and
Localization: An Improved \$\textbackslash widehat\{R\}\$ for Assessing
Convergence of MCMC,'' March. \url{http://arxiv.org/abs/1903.08008}.

\leavevmode\hypertarget{ref-vinknd}{}%
Vink, Gerko. n.d. ``Towards a Standardized Evaluation of Multiple
Imputation Routines.''

\leavevmode\hypertarget{ref-vink14}{}%
Vink, Gerko, and Stef van Buuren. 2014. ``Pooling Multiple Imputations
When the Sample Happens to Be the Population.'' \emph{arXiv:1409.8542
{[}Math, Stat{]}}, September. \url{http://arxiv.org/abs/1409.8542}.

>>>>

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\begin{document}

\title{Missing the Point\(\colon\) Non-Convergence in Multiple Imputation
Procedures}

\runninghead{Oberman}

\author{H. I. Oberman\affilnum{1}}

\affiliation{\affilnum{1}{Department of Methodology and Statistics, Utrecht University, Utrecht,
The Netherlands}}

\corrauth{Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht,
The Netherlands.}

\email{\href{mailto:h.i.oberman@uu.nl}{\nolinkurl{h.i.oberman@uu.nl}}}

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Aenean ut elit
odio. Donec fermentum tellus neque, vitae fringilla orci pretium vitae.
Fusce maximus finibus facilisis. Donec ut ullamcorper turpis. Donec ut
porta ipsum. Nullam cursus mauris a sapien ornare pulvinar. Aenean
malesuada molestie erat quis mattis. Praesent scelerisque posuere
faucibus. Praesent nunc nulla, ullamcorper ut ullamcorper sed, molestie
ut est. Donec consequat libero nisi, non semper velit vulputate et.
Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget.
Curabitur aliquet vehicula quam non pulvinar. Aliquam facilisis tortor
nec purus finibus, sit amet elementum eros sodales. Ut porta porttitor
vestibulum. Integer molestie, leo ut maximus aliquam, velit dui iaculis
nibh, eget hendrerit purus risus sit amet dolor. Sed sed tincidunt ex.
Curabitur imperdiet egestas tellus in iaculis. Maecenas ante neque,
pretium vel nisl at, lobortis lacinia neque. In gravida elit vel
volutpat imperdiet. Sed ut nulla arcu. Proin blandit interdum ex sit
amet laoreet. Phasellus efficitur, sem hendrerit mattis dapibus, nunc
tellus ornare nisi, nec eleifend enim nibh ac ipsum. Aenean tincidunt
nisl sit amet facilisis faucibus. Donec odio erat, bibendum eu imperdiet
sed, gravida luctus turpis.
\end{abstract}

\keywords{MICE; convergence;}

\maketitle

@@

At some point, any scientist conducting statistical analyses will run
into a missing data problem \citep{alli02}. Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results \citep{buur18}. A popular answer
to the ubiquitous problem of missing information is to use the framework
of multiple imputation (MI), proposed by \citet{rubin87}. For the
theoretical foundation of MI, see \citet{rubin87}. For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model}. The imputation model can be a
joint model for all variables, or several univariate conditional
distributions (fully conditional specification). FCS is more flexible
and ``avoids the problem of specifying an appropriate joint imputation
distribution and replaces this by the selection of appropriate
univariate conditional distributions''. With FCS, an iterative
algorithmic procedure is used to \ldots{} multiple imputation by chained
equations \ldots{} ``The values for imputing the missing data are drawn
from the posterior predictive distribution of the missing data
conditional on the observed data.'' There is no scientific consensus on
the convergence properties of FCS algorithms \citep{taka17}. Some
default techniques in \texttt{mice} might not yield converged states at
all \citep{murr18}. Therefore, algorithmic convergence should be
monitored carefully.

Convergence cannot be determined for sure, because MICE is an MCMC
method, and convergence of MCMC is not from scalar to a point, but from
one distribution to another. Generated values will vary even after
convergence. Therefore we can only diagnose non-convergence, not
convergence \citep{hoff09}. ``a weak diagnostics is better than no
diagnostic at all'' \citep{cowl96}.

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye
\citep[\(\S\) 6.5.2]{buur18}. Therefore diagnostic evaluation of
convergence would be preferred. However, not known whether MCMC
diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied \citep{buur18}. How can non-convergence be diagnosed? How many
iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 1;
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

For reasons of brevity, we only focus on the MI algorithm implemented in
the world-leading MI software: the \texttt{R} \citep{R} package
\texttt{mice} \citep{mice}. The convergence properties of this MI
algorithm are investigated through model-based simulation. The results
of this simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets.

@@

The stationarity component of convergence may be evaluated with
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending
within chains.

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) \citep[a.k.a. `Gelman-Rubin
statistic';][]{gelm92}. With an adapted version of \(\widehat{R}\),
proposed by \citet{veht19}, we might also evaluate the stationarity
component of convergence. This would make \(\widehat{R}\) a general
convergence diagnostic. The application of \(\widehat{R}\) to assess
stationarity has not been thoroughly investigated. Therefore, this study
employs both \(\widehat{R}\) and autocorrelation to investigate
convergence, as recommended by \citep[p.~898]{cowl96}.

Note that all of these methods evaluate the convergence of univariate
scalar summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). \citet{buur18} proposed to
implement multivariable evaluation through eigenvalue decomposition
\citep{mack03}. This method is outside of the scope of the current study
(\textbf{not anymore!}).

@@

To define \(\widehat{R}\), we follow notation by \citep[p.~5]{veht19}.
Let \(M\) be the total number of chains, \(T\) the number of iterations
per chain, and \(\theta\) the scalar summary of interest (e.g., chain
mean or chain variance). For each chain (\(m = 1, 2, \dots, M\)), we
estimate the variance of \(\theta\), and average these to obtain
within-chain variance \(W\).

@@

We can interpret \(\widehat{R}\) as potential scale reduction factor
since it indicates by how much the variance of \(\theta\) could be
shrunken down if an infinite number of iterations per chain would be run
\citep{gelm92}. This interpretation assumes that chains are
`over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
Over-dispersion implies that the initial values of the chains are `far
away' from the target distribution and each other. When all chains
sample independent of their initial values, the mixing component of
convergence is satisfied, and \(\widehat{R}\)-values will be close to
one. High \(\widehat{R}\)-values thus indicate non-convergence. The
conventionally acceptable threshold for convergence was
\(\widehat{R} < 1.2\) \citep{gelm92}. More recently, \citet{veht19}
proposed a more stringent threshold of \(\widehat{R} < 1.01\).

@@

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain \citep[p.~147]{lync07}. In this study we only consider \(AC\) at
lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\)
iteration of the same chain.

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. That is, there is no baseline measure available to
evaluate performance against.The current practice of visually inspecting
imputation chains for signs of non-mixing or non-stationarity can only
diagnose severely pathological cases of non-convergence. We could also
look at distributional characteristics, and plausibility of imputed
values, see \citet{vinknd}. For now, this is outside of the scope of
this study.

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). No hypothesis
was formulated about the performance of \(AC\) as convergence
diagnostic.

@@

The convergence of \texttt{mice} is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

\begin{verbatim}
# pseudo-code of simulation 
simulate data 
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    create missingness
    impute the missingness
    compute convergence diagnostics
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
\end{verbatim}

@@

\(\widehat{R}\) is computed by implementing Vehtari et al.'s
\citeyearpar{veht19} recommendations. \(AC\) is computed with function
\texttt{stats::acf()}.

To estimate the quantity of scientific interest, \(Q\), we perform
multiple linear regression on each completed dataset with the function
\texttt{stats::lm()}. We obtain an estimated regression coefficient per
imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
the function \texttt{mice::pool()} to get variance estimates according
to Rubin's \citeyearpar{rubin87} rules, and subsequently implement
finite population pooling conform \citet{vink14}.

@@

\textbf{Add more info about figure legends and axes.}

\begin{verbatim}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{Convergence diagnostics (means)}
\end{figure}

\begin{verbatim}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Convergence diagnostics (variances)}
\end{figure}

\begin{verbatim}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Convergence diagnostics (regression coefficients)}
\end{figure}

@@

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% \citep{neym34}, namely 95\%. Similar
to the CIWs, there is some trending across iterations (i.e., the Loess
line is never flat).

@@

\bibliographystyle{sageh}
\bibliography{thesis.bib}
